---
title: Bio-logging Citation Network
format:
  html:
    code-fold: true
---

On 2023-02-10, I searched PubMed for `"bio-logg*" OR biologg* OR biotelemetry`, returning 970 results. I exported the results to data/pubmed-naive.csv. Manually pruning unrelated studies (e.g., biomedical or other life sciences) left 501 bio-logging studies.

Now I'll try to use `rcrossref` to get the works cited by the first publication.

```{r}
#| label: setup
#| include: false

library(janitor)
library(rcrossref)
library(tidyverse)
```

```{r}
#| label: read-pruned

pubmed <- readxl::read_excel(here::here("data", "pubmed-pruned.xlsx"),
                             sheet = 1) %>% 
  clean_names() %>% 
  filter(retained == "y") %>% 
  arrange(order)

```

```{r}
#| label: fetch-one

one_work <- cr_works(dois = pubmed$doi[1])
one_refs <- one_work$data$reference[[1]]
one_refs

```

Now all the works cited.

```{r}
#| label: fetch-refs

pubmed_cited_path <- here::here("data", "pubmed_cited.rds")

if (!file.exists(pubmed_cited_path)) {
  works <- cr_works(dois = pubmed$doi)
  pubmed_cited <- map2_dfr(works$data$doi, 
                           works$data$reference, 
                           function(work, refs) {
                             tibble(
                               citing_doi = work,
                               work_cited = refs$DOI
                             )
                           }) %>% 
    drop_na()
  
  saveRDS(pubmed_cited, pubmed_cited_path)
} else {
  pubmed_cited <- readRDS(pubmed_cited_path)
}

```

What are the most cited? Red points indicate paper is in the original PubMed results.

```{r}
#| label: most-cited

cite_count <- pubmed_cited %>% 
  mutate(work_cited = str_to_lower(work_cited)) %>% 
  count(work_cited) %>% 
  arrange(desc(n)) %>% 
  mutate(is_pubmed = map_lgl(work_cited, ~ .x %in% pubmed$doi))

with_rank <- cite_count %>% 
  mutate(rank = row_number()) %>% 
  slice(1:500)

ggplot(with_rank, aes(rank, n)) +
  geom_line(color = "grey") +
  geom_point(data = filter(with_rank, is_pubmed), color = "red") +
  labs(x = "Citation rank",
       y = "Citations from PubMed query") +
  theme_minimal()
```

Fetch metadata for papers not in original query cited by 5+ papers in the original query (634 papers).

```{r}
#| label: cited-meta

pubmed_cited_full_path <- here::here("data", "pubmed_cited_full.rds")

if (!file.exists(pubmed_cited_full_path)) {
  cited_dois <- with(cite_count, work_cited[n >= 5 & !is_pubmed])
  pubmed_cited_full <- cr_works(dois = cited_dois)$data
  saveRDS(pubmed_cited_full, pubmed_cited_full_path)
} else {
  pubmed_cited_full <- readRDS(pubmed_cited_full_path)
}

get_first_author <- function(author_col) {
  map_chr(author_col,
          function(author) {
            result <- tryCatch(
              author$family[1],
              error = function(e) ""
            )
            if (!is.character(result)) result <- ""
            result
          })
}
get_all_authors <- function(author_col) {
  map_chr(author_col,
          function(author) {
            result <- tryCatch(
              paste(paste(author$given, author$family), collapse = ", "),
              error = function(e) ""
            )
            if (!is.character(result)) result <- ""
            result
          })
}

# Save to CSV
pubmed_cited_full %>% 
  transmute(container.title,
            created,
            doi,
            title,
            type,
            first.author = get_first_author(author),
            authors = get_all_authors(author)) %>%
  write_csv(here::here("data", "pubmed_cited_full.csv"))

# display top 10 most cited
pubmed_cited_full %>% 
  left_join(mutate(cite_count, work_cited = str_to_lower(work_cited)), 
            by = c(doi = "work_cited")) %>% 
  mutate(first.author = str_to_sentence(get_first_author(author))) %>% 
  transmute(citation = str_glue("{first.author} ({str_sub(created, 1, 4)})"),
            container = str_sub(container.title, 1, 20),
            title = str_sub(str_to_sentence(title), 1, 20),
            citations = n) %>% 
  slice(1:10)

```
